---
title: "Impact of Rebalancing Imbalanced Data"
author: "Govind Nair"
date: "3/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This post looks at the impact of rebalancing an imbalanced dataset and seeks to identify some heuristics/rules of thumb to determine when it is appropriate to rebalance an imabalanced dataset through oversampling.



## Case 1: When your model is a good fit for your data

### Original Data
Consider the true data you are working with below which appears to be linearly seperable. The data is imbalanced with a ratio of 1:4.

```{r, echo=FALSE}
set.seed(0)
c1 <- cbind('x1'=rnorm(200,0.6,0.12),'x2'=rnorm(200,0.4,0.15))
c2 <- cbind('x1'=rnorm(800,0.4,0.12),'x2'=rnorm(800,0.6,0.15))
original_data <- data.frame(rbind(c1,c2))
original_data$y <- as.factor(c(rep(1,200),rep(0,800)))
```

The data and the decision boundary from the simple logistic regression model built on the data is shown below.
The decision boundary corresponds to a contour where the proability of the minority class = 0.2.

```{r}
glm1 <- glm(y ~ ., data = original_data, family = 'binomial')
summary(glm1)
```


The 95% confidence intervals and the data to plot decision boundaries are created.

```{r calc_conf_intervals1}
grid_dimension <- seq(0,1,by =0.01)
grid <- expand.grid(x1=grid_dimension,x2 = grid_dimension)
grid_preds <- predict(glm1,newdata=grid,type='response')

preds_ci <-  predict(glm1,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties
upr2 <- glm1$family$linkinv(upr)
lwr2 <- glm1$family$linkinv(lwr)

contour_data <- cbind(grid,lwr = lwr2,upr =upr2,pred = grid_preds)
```


The plot the data and the decision boundary along with the 95% prediction intervals are shown below.


```{r original_data_plot}


library(ggplot2)
p <- ggplot(original_data,aes(x=x1,y=x2))+geom_point(aes(colour = factor(y)),alpha =0.3)+labs(title='Original Data') + theme (legend.title = element_blank())
p


p+ stat_contour(data = contour_data,aes(x = x1,y = x2,z=pred),breaks=c(0.2))+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=lwr),breaks=c(0.2),linetype=2)+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=upr),breaks=c(0.2),linetype=2)


```


### Oversampled Data

Now we oversample the minority class.

```{r }

minority_class <- subset(original_data,original_data$y==1)
majority_class <- subset(original_data,original_data$y==0)
oversamples <- minority_class[sample(c(1:nrow(minority_class)),600,replace = TRUE),]
oversampled_data <- rbind(oversamples,minority_class,majority_class)
```

A second model is fitted on the oversampled data.
```{r}
glm2 <- glm(y ~ ., data = oversampled_data, family = 'binomial')
summary(glm2)

```

The plot is shown below with a 95% confidence limits.

```{r oversampled_data_plot}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm2,newdata=grid,type='response')

preds_ci <-  predict(glm2,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties

upr2 <- glm3$family$linkinv(upr)
lwr2 <-glm3$family$linkinv(lwr)

contour_data <- cbind(grid,lwr = lwr2,upr =upr2,pred = grid_preds)


p <- ggplot(oversampled_data,aes(x=x1,y=x2))+geom_point(aes(colour = factor(y)),alpha =0.3)+labs(title='Oversampled Data') + theme (legend.title = element_blank())



p+ stat_contour(data = contour_data,aes(x = x1,y = x2,z=pred),breaks=c(0.2))+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=lwr),breaks=c(0.2),linetype=2)+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=upr),breaks=c(0.2),linetype=2)
```

### Undersampled data

Now we try undersampling the majority class. This is typically done only when the data is really big and the tools/infrastructure being used does not support really big data.


```{r undersampled_data_plot}
undersamples <- majority_class[sample(c(1:nrow(majority_class)),200,replace = TRUE),]
undersampled_data <- rbind(minority_class,undersamples)

```


A third model is fitted on the undersampled data.
```{r}
glm3 <- glm(y ~ ., data = undersampled_data, family = 'binomial')
summary(glm3)

```

The plot is shown below with a 95% confidence limits.

```{r}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm3,newdata=grid,type='response')

preds_ci <-  predict(glm3,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties

upr2 <- glm3$family$linkinv(upr)
lwr2 <-glm3$family$linkinv(lwr)

contour_data <- cbind(grid,lwr = lwr2,upr =upr2,pred = grid_preds)


p <- ggplot(undersampled_data,aes(x=x1,y=x2))+geom_point(aes(colour = factor(y)),alpha =0.3)+labs(title='Undersampled Data') + theme (legend.title = element_blank())



p+ stat_contour(data = contour_data,aes(x = x1,y = x2,z=pred),breaks=c(0.2))+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=lwr),breaks=c(0.2),linetype=2)+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=upr),breaks=c(0.2),linetype=2)
```
## Case 2: When your model is not a good fit for your data

Now consider the case when the model is not a good fit for the data.

### Original Data

```{r}
set.seed(0)
#class 1
c1 <- cbind('x1'=c(rnorm(100,0.2,0.1),rnorm(100,0.8,0.1)),
            'x2'=c(rnorm(100,0.8,0.1),rnorm(100,0.2,0.1)))
#class2
c2 <- cbind('x1'=rnorm(800,0.5,0.1),'x2'=rnorm(800,0.5,0.1))

original_data2 <- data.frame(rbind(c1,c2))
original_data2$y <- as.factor(c(rep(1,200),rep(0,800)))

```


A model is fitted on the original data.
```{r}
glm4 <- glm(y ~ ., data = original_data2, family = 'binomial')
summary(glm4)

```


The plot is shown below

```{r original_data_plot2,warning=FALSE,message=FALSE}
grid_preds <- predict(glm4,newdata=grid,type='response')

preds_ci <-  predict(glm4,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties

upr2 <- glm4$family$linkinv(upr)
lwr2 <-glm4$family$linkinv(lwr)

contour_data <- cbind(grid,lwr = lwr2,upr =upr2,pred = grid_preds)


p <- ggplot(original_data2,aes(x=x1,y=x2))+geom_point(aes(colour = factor(y)),alpha =0.3)+labs(title='Original Data') + theme (legend.title = element_blank())



p+ stat_contour(data = contour_data,aes(x = x1,y = x2,z=pred),breaks=c(0.2))+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=lwr),breaks=c(0.2),linetype=2)+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=upr),breaks=c(0.2),linetype=2)

```


### Oversampled Data

Now we oversample the minority class.

```{r}

minority_class <- subset(original_data2,original_data2$y==1)
majority_class <- subset(original_data2,original_data2$y==0)
oversamples <- minority_class[sample(c(1:nrow(minority_class)),600,replace = TRUE),]
oversampled_data2 <- rbind(oversamples,minority_class,majority_class)
```

A second model is fitted on the oversampled data.
```{r}
glm5 <- glm(y ~ ., data = oversampled_data2, family = 'binomial')
summary(glm5)

```


The plot is shown below

```{r oversampeld_data_plot2, warning=FALSE,message=FALSE}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm5,newdata=grid,type='response')

preds_ci <-  predict(glm5,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties

upr2 <- glm5$family$linkinv(upr)
lwr2 <-glm5$family$linkinv(lwr)

contour_data <- cbind(grid,lwr = lwr2,upr =upr2,pred = grid_preds)


p <- ggplot(oversampled_data2,aes(x=x1,y=x2))+geom_point(aes(colour = factor(y)),alpha =0.3)+labs(title='Oversampled Data') + theme (legend.title = element_blank())



p+ stat_contour(data = contour_data,aes(x = x1,y = x2,z=pred),breaks=c(0.5))+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=lwr),breaks=c(0.5),linetype=2)+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=upr),breaks=c(0.5),linetype=2)
```


### Undersampled Data

Now we oversample the minority class.



```{r }
undersamples <- majority_class[sample(c(1:nrow(majority_class)),200,replace = TRUE),]
undersampled_data2 <- rbind(minority_class,undersamples)

```


A second model is fitted on the oversampled data.
```{r}
glm6 <- glm(y ~ ., data = undersampled_data2, family = 'binomial')
summary(glm6)

```


The plot is shown below

```{r undersampled_data_plot2, warning=FALSE,message=FALSE}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm6,newdata=grid,type='response')

preds_ci <-  predict(glm6,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties

upr2 <- glm6$family$linkinv(upr)
lwr2 <-glm6$family$linkinv(lwr)

contour_data <- cbind(grid,lwr = lwr2,upr =upr2,pred = grid_preds)


p <- ggplot(undersampled_data2,aes(x=x1,y=x2))+geom_point(aes(colour = factor(y)),alpha =0.3)+labs(title='UnderSampled Data') + theme (legend.title = element_blank())



p+ stat_contour(data = contour_data,aes(x = x1,y = x2,z=pred),breaks=c(0.5))+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=lwr),breaks=c(0.5),linetype=2)+
        stat_contour(data = contour_data,aes(x = x1,y = x2,z=upr),breaks=c(0.5),linetype=2)
```