---
title: "Impact of Rebalancing Imbalanced Data"
author: "Govind Nair"
date: "3/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This post looks at the impact of rebalancing an imbalanced dataset and seeks to identify some heuristics/rules of thumb to determine when it is appropriate to rebalance an imabalanced dataset through oversampling/undersampling.




## Case 1: When your model is a good fit for your data

### Original Data
Consider the true data you are working with below which appears to be linearly seperable. The data is imbalanced with a ratio of 1:4.

```{r, echo=FALSE}
set.seed(0)
c1 <- cbind('x1'=rnorm(20,0.6,0.12),'x2'=rnorm(20,0.4,0.15))
c2 <- cbind('x1'=rnorm(80,0.4,0.12),'x2'=rnorm(80,0.6,0.15))
original_data <- data.frame(rbind(c1,c2))
original_data$y <- as.factor(c(rep(1,20),rep(0,80)))
```

The data and the decision boundary from the simple logistic regression model built on the data is shown below.
The decision boundary corresponds to a contour where the porbability of the minority class = 0.2.

```{r}
glm1 <- glm(y ~ ., data = original_data, family = 'binomial')
summary(glm1)
```


The plot the data and the deicision boundary along with the 95% prediction intervals are shown below.

```{r}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm1,newdata=grid,type='response')

preds_ci <-  predict(glm1,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties
upr2 <- glm1$family$linkinv(upr)
lwr2 <- glm1$family$linkinv(lwr)



nrow <- length(seq(0,1,by=0.01))
plot(original_data$x1,original_data$x2,col=original_data$y,pch = 16,xlab='x1',ylab='x2',main = 'Original Data')
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(grid_preds,nrow =nrow),levels=0.2,
        col="cornflowerblue",lwd=2,drawlabels=FALSE,add=TRUE)
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(upr2,nrow =nrow),levels=0.2,
        col="green",lwd=2,drawlabels=FALSE,add=TRUE,lty=2)
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(lwr2,nrow =nrow),levels=0.2,
        col="green",lwd=2,drawlabels=FALSE,add = TRUE,lty=2)

```


### Oversampled Data

Now we oversample the minority class.

```{r}

minority_class <- subset(original_data,original_data$y==1)
majority_class <- subset(original_data,original_data$y==0)
oversamples <- minority_class[sample(c(1:nrow(minority_class)),60,replace = TRUE),]
oversampled_data <- rbind(oversamples,minority_class,majority_class)
```

A second model is fitted on the oversampled data.
```{r}
glm2 <- glm(y ~ ., data = oversampled_data, family = 'binomial')
summary(glm2)

```

The plot is shown below with a 95% confidence limits.

```{r}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm2,newdata=grid,type='response')

preds_ci <-  predict(glm2,newdata=grid,type='link',se.fit=TRUE)

#Calculate 95% prediction interval bounds
critval <- 1.96 ## approx 95% CI
upr <- preds_ci$fit + (critval * preds_ci$se.fit)
lwr <- preds_ci$fit - (critval * preds_ci$se.fit)



#These are logits, use inverse link function to get probabilties

upr2 <- glm1$family$linkinv(upr)
lwr2 <-glm1$family$linkinv(lwr)

nrow <- length(seq(0,1,by=0.01))
plot(oversampled_data$x1,oversampled_data$x2,col=oversampled_data$y,pch = 16,xlab='x1',ylab='x2',main='Oversampled data')
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(grid_preds,nrow =nrow),levels=0.5,
        col="cornflowerblue",lwd=2,drawlabels=FALSE,add=TRUE)
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(upr2,nrow =nrow),levels=0.5,
        col="green",lwd=2,drawlabels=FALSE,add=TRUE,lty=2)
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(lwr2,nrow =nrow),levels=0.5,
        col="green",lwd=2,drawlabels=FALSE,add = TRUE,lty=2)
```

## Case 2: When your model is not a good fit for your data

Now consider the case when the model is not a good fit for the data.

### Original Data

```{r}
set.seed(0)
c1 <- cbind('x1'=c(rnorm(10,0.6,0.15),rnorm(10,0.2,0.15)),'x2'=c(rnorm(10,0.4,0.15),rnorm(10,0.8,0.15)))
c2 <- cbind('x1'=rnorm(80,0.4,0.15),'x2'=rnorm(80,0.6,0.15))

original_data2 <- data.frame(rbind(c1,c2))
original_data2$y <- as.factor(c(rep(1,20),rep(0,80)))

```


A model is fitted on the original data.
```{r}
glm3 <- glm(y ~ ., data = original_data2, family = 'binomial')
summary(glm3)

```


The plot is shown below

```{r}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm3,newdata=grid,type='response')



nrow <- length(seq(0,1,by=0.01))
plot(original_data2$x1,original_data2$x2,col=original_data2$y,pch = 16,xlab='x1',ylab='x2',main='Original Data')
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(grid_preds,nrow =nrow),levels=0.2,
        col="cornflowerblue",lwd=2,drawlabels=FALSE,add=TRUE)

```


### Oversampled Data

Now we oversample the minority class.

```{r}

minority_class <- subset(original_data2,original_data2$y==1)
majority_class <- subset(original_data2,original_data2$y==0)
oversamples <- minority_class[sample(c(1:nrow(minority_class)),60,replace = TRUE),]
oversampled_data2 <- rbind(oversamples,minority_class,majority_class)
```

A second model is fitted on the oversampled data.
```{r}
glm4 <- glm(y ~ ., data = oversampled_data2, family = 'binomial')
summary(glm4)

```


The plot is shown below

```{r}
grid <- expand.grid(x1=seq(0,1,by=0.01),x2 = seq(0,1,by=0.01))
grid_preds <- predict(glm4,newdata=grid,type='response')
nrow <- length(seq(0,1,by=0.01))
contour(x=seq(0,1,by=0.01), y=seq(0,1,by=0.01), z=matrix(grid_preds,nrow =nrow),levels=0.5,
        col="cornflowerblue",lwd=2,drawlabels=FALSE)
points(oversampled_data2$x1,oversampled_data2$x2,col=oversampled_data2$y,pch = 16)
```
